---
title: "example1"
output: pdf_document
date: "2025-02-04"
---

## Simulate data

```{r}
# taken from
# https://github.com/stephenslab/drift.alpha/blob/master/R/simulate.R
two_pop_tree_sim <- function(n_per_pop, p, sigma_e, sigma_b=rep(1, 3)){

  if(length(sigma_b) != 3){
    stop("There can only be 3 branch lengths i.e. factors")
  }

  # number of individuals
  n <- n_per_pop * 2

  # loadings matrix which specifies the topology
  L <- matrix(0, nrow=n, ncol=3)
  L[ ,1] <- 1
  L[1:n_per_pop, 2] <- 1
  L[(n_per_pop + 1):(2 * n_per_pop), 3] <- 1

  # drift on each branch
  F <- matrix(NA, nrow=p, ncol=3)
  for(k in 1:3){
    F[ ,k] <- rnorm(p, 0, sigma_b[k])
  }

  # errors
  if (length(sigma_e) == 1) {
    E <- matrix(rnorm(n*p, 0, sigma_e), nrow=n, ncol=p)
  } else if (length(sigma_e) == n) {
    E <- matrix(0, nrow=n, ncol=p)
    for (i in 1:n) {
      E[i,] <- rnorm(p, 0, sigma_e[i])
    }
  } else {
    E <- matrix(0, nrow=n, ncol=p)
    for (j in 1:p) {
      E[,j] <- rnorm(n, 0, sigma_e[j])
    }
  }

  # data
  Y <- L %*% t(F) + E

  # simulation object
  res <- list(Y=Y, L=L, F=F)
  return(res)

}
```

## Var type = 0

```{r}
set.seed(235)
n_per_pop <- 50
pops <- c(rep("Pop1", n_per_pop), rep("Pop2", n_per_pop))
sigma_e <- 1.0
sigma_b <- c(1.0, 1.0, 1.0)
p = 50
sim_res <- two_pop_tree_sim(n_per_pop, p, sigma_e, sigma_b)
```


### EBMF with joint update on F with normal priors

```{r}
library(flashier)
library(devtools)
load_all()
library(flashierX)
flash_obj <- flash(sim_res$Y, ebnm_fn = c(ebnm_point_laplace, ebnm_normal), greedy_Kmax = 10, var_type = 0)
cat('Initial elbo: ', flash_obj$elbo, '\n')
flash_obj <- flash_backfitX(flash_obj)
```

### Regular EBMF with sequential update on F with normal priors

```{r}
flash_obj_old <- flash(sim_res$Y, ebnm_fn = c(ebnm_point_laplace, ebnm_normal), greedy_Kmax = 10, backfit = T, var_type = 0)
flash_obj_old$elbo
```

### Comparison

```{r}
plot(flash_obj$elbo.df)
norm(sim_res$Y %*% t(sim_res$Y) / 100 - flash_obj$L_pm %*% t(flash_obj$L_pm))
norm(sim_res$Y %*% t(sim_res$Y) / 100 - flash_obj_old$L_pm %*% t(flash_obj_old$L_pm))
norm(sim_res$Y - flash_obj$L_pm %*% t(flash_obj$F_pm))
norm(sim_res$Y - flash_obj_old$L_pm %*% t(flash_obj_old$F_pm))
heatmap(flash_obj$L_pm, Rowv = NA, Colv = NA, scale = "column", col = heat.colors(256), margins = c(5, 5))
heatmap(flash_obj_old$L_pm, Rowv = NA, Colv = NA, scale = "column", col = heat.colors(256), margins = c(5, 5))
```

## Var type = 1

```{r}
set.seed(235)
n_per_pop <- 50
pops <- c(rep("Pop1", n_per_pop), rep("Pop2", n_per_pop))
sigma_e <- rep((1:4) / 4, 25)
sigma_b <- c(1.0, 1.0, 1.0)
p = 50
sim_res <- two_pop_tree_sim(n_per_pop, p, sigma_e, sigma_b)
```


### EBMF with joint update on F with normal priors

```{r}
library(flashier)
library(devtools)
load_all()
library(flashierX)
flash_obj <- flash(sim_res$Y, ebnm_fn = c(ebnm_point_laplace, ebnm_normal), greedy_Kmax = 10, var_type = 1)
cat('Initial elbo: ', flash_obj$elbo, '\n')
flash_obj <- flash_backfitX(flash_obj)
```

### Regular EBMF with sequential update on F with normal priors

```{r}
flash_obj_old <- flash(sim_res$Y, ebnm_fn = c(ebnm_point_laplace, ebnm_normal), greedy_Kmax = 10, backfit = T, var_type = 1)
flash_obj_old$elbo
```

### Comparison

```{r}
plot(flash_obj$elbo.df)
norm(sim_res$Y %*% t(sim_res$Y) / 100 - flash_obj$L_pm %*% t(flash_obj$L_pm))
norm(sim_res$Y %*% t(sim_res$Y) / 100 - flash_obj_old$L_pm %*% t(flash_obj_old$L_pm))
norm(sim_res$Y - flash_obj$L_pm %*% t(flash_obj$F_pm))
norm(sim_res$Y - flash_obj_old$L_pm %*% t(flash_obj_old$F_pm))
heatmap(flash_obj$L_pm, Rowv = NA, Colv = NA, scale = "column", col = heat.colors(256), margins = c(5, 5))
heatmap(flash_obj_old$L_pm, Rowv = NA, Colv = NA, scale = "column", col = heat.colors(256), margins = c(5, 5))
```

## Var type = 2

```{r}
set.seed(235)
n_per_pop <- 50
pops <- c(rep("Pop1", n_per_pop), rep("Pop2", n_per_pop))
sigma_e <- rep((1:5) / 5, 10)
sigma_b <- c(1.0, 1.0, 1.0)
p = 50
sim_res <- two_pop_tree_sim(n_per_pop, p, sigma_e, sigma_b)
```


### EBMF with joint update on F with normal priors

```{r}
library(flashier)
library(devtools)
load_all()
library(flashierX)
flash_obj <- flash(sim_res$Y, ebnm_fn = c(ebnm_point_laplace, ebnm_normal), greedy_Kmax = 10, var_type = 2)
cat('Initial elbo: ', flash_obj$elbo, '\n')
flash_obj <- flash_backfitX(flash_obj)
```

### Regular EBMF with sequential update on F with normal priors

```{r}
flash_obj_old <- flash(sim_res$Y, ebnm_fn = c(ebnm_point_laplace, ebnm_normal), greedy_Kmax = 10, backfit = T, var_type = 2)
flash_obj_old$elbo
```

### Comparison

```{r}
plot(flash_obj$elbo.df)
norm(sim_res$Y %*% t(sim_res$Y) / 100 - flash_obj$L_pm %*% t(flash_obj$L_pm))
norm(sim_res$Y %*% t(sim_res$Y) / 100 - flash_obj_old$L_pm %*% t(flash_obj_old$L_pm))
norm(sim_res$Y - flash_obj$L_pm %*% t(flash_obj$F_pm))
norm(sim_res$Y - flash_obj_old$L_pm %*% t(flash_obj_old$F_pm))
heatmap(flash_obj$L_pm, Rowv = NA, Colv = NA, scale = "column", col = heat.colors(256), margins = c(5, 5))
heatmap(flash_obj_old$L_pm, Rowv = NA, Colv = NA, scale = "column", col = heat.colors(256), margins = c(5, 5))
```